---
title: 'risa: Basic functions'
author: "Joao Bosco Gusmao"
date: "2026-04-02"
output:
  pdf_document: default
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

## The *risa* package

The **risa** package, which stands for RISk Assessment, was designed to
automate the applycation of the Habitat Risk Assessment (HRA) model,
which is part of the *InVEST* (Integrated Valuation of Ecosystem
Services and Tradeoffs) toolkit developed by the Natural Capital Project
(Arkema et al., 2014; Sharp et al., 2020).

There are a entire suite of functions that were projected to automate
the steps to apply the *ByRA* (Bycatch Risk Assessment) toolkit (Hines et al., 2020; Wilhelm
et al., 2022). *ByRA* uses HRA methodology to quantify the spatial
overlap between fisheries and marine megafauna and has been applied to
assess bycatch risk for marine mammals.

Specifically, **risa** automates the generation of Kernel Density Maps
of species occurrences and stressor intensities from either data frames
or shapefiles, the creation of overlap hotspot maps, and the execution
of the full HRA model for one or multiple species and stressors. Users
can select among multiplicative and Euclidean formulations of risk,
apply optional decay functions to account for stressor influence
buffers, and generate detailed summary statistics. In addition, **risa**
provides flexibility to customize model settings, including the number
of risk classes, buffer sizes for kernel density estimates and stressor
zones, and the choice of coordinate system (metric or decimal).

## Installation

Since **risa** is still in its *alpha* version, it cannot be installed
from R's CRAN repository. Thus, **risa** must be installed from a GitHub
repository, which can be assessed via the link:

<https://github.com/GusmaoJB/risa/>

The easiest way to install **risa** is using the *install_github()* function in the package 
**devtools**. If you don't have it, install **devtools** and via *install.packages("devtools")* and then run the code below to install **risa**.

```{r message=FALSE, warning=FALSE, eval=FALSE}
# Load devtools
library(devtools)

# install the risa package
devtools::install_github("GusmaoJB/risa")

# Load the risa package
library(risa)
```

## Performing ByRA

Since we depend on functions in the packages *terra* and *sf* to load
and manipulate rasters and shapefiles, we will load these packages to
perform our examples. First, let's keep our work files (rasters,
shapefiles, dataframes) in a *data* folder inside our working directory.
In this first example, let's suppose we have species and stressor
distribuions as occurrence points in CSV files:

```{r warning=FALSE, include=FALSE}
library(risa)
spp_df <- read.csv("/home/jojo/Documents/pontal_projects/risa2/tests/data/species_distr.csv")
str_df <- read.csv("/home/jojo/Documents/pontal_projects/risa2/tests/data/stressor_distr.csv")
criteria <- read.csv("/home/jojo/Documents/pontal_projects/risa2/tests/data/criteria.csv")
```

```{r message=FALSE, warning=FALSE, eval=FALSE}
# Load important packages
library(sf)
library(terra)

# Loading species and stressor data
spp_df <- read.csv("data/species_distr.csv")
str_df <- read.csv("data/stressor_distr.csv")
```

Then we have a dataframe describing the occurrences of different
species:

```{r}
# Check the first rows
head(spp_df)
```

And a data.frame with similar structure for stressors:

```{r}
# Check the first rows
head(str_df)
```

All we need now is to load the criteria table. This table compiles the key information required for the analysis, including exposure ratings of each species (or habitat) to specific stressors, consequence ratings of those interactions, and the corresponding weights used in risk estimation and data quality assessment.

Since we are working within the **ByRA** framework, we will use the criteria table specifically designed for this purpose, following the approach described in Hines et al. (2020).

```{r message=FALSE, warning=FALSE, eval=FALSE}
# Load criteria data
criteria <- read.csv("data/criteria.csv")

# Check the first rows
head(criteria)
```

```{r echo=FALSE, warning=FALSE}
head(criteria)
```

Now we already have everything we need to run our first ByRA!

Let’s start with a very simple example, where we just want a quick assessment of bycatch risk for two megafaunal species exposed to two different fishing gears. Since this is a first try, we will keep things straightforward and rely on the default behavior of the quick_byra() function.

This means we will not worry (yet) about the details behind how kernel density estimates for species and stressors are constructed, or how the HRA model is calculated. The default settings handle all of that for us, using conservative assumptions and standard data preparation protocols typically recommended for ByRA analyses.

We can always refine and tweak both the data prep and the HRA model later—so for now, let’s start simple and see what we get.

```{r}
byra1 <- quick_byra(spp_df, str_df, criteria)
```

And that is it! All the results we want, including the maps and statistics of risk estimates 
are stored in *byra1*. If we check it out, we will find this:

```{r}
names(byra1)
```

Which means:

- **Dolphin and Turtle:** Partial results (risk estimates) for each species (can be just one or more!). This is a list containing maps of the Exposure and Consequence components of the risk assessment, as well as overall risk estimates for each individual stressor and their combined effects. It also includes a table with summary statistics for these estimates.

- **ecosys_risk_raw:** A `SpatRaster` map of the raw ecosystem risk estimates, representing the combined risk across all species considered in the assessment (Dolphin and Turtle).

- **ecosys_risk_classified:** A `SpatRaster` map of reclassified ecosystem risk estimates. ByRA typically uses four risk categories: 0 = None, 1 = Low, 2 = Medium, and 3 = High.

- **summary_stats:** A `data.frame` with summary statistics of risk estimates at both the ecosystem and species levels. It includes results for individual and combined stressor effects, as well as minimum, maximum, and mean values of Exposure, Consequence, and risk estimates, in addition to the relative area of each risk category.

- **kde_maps:** A list containing all kernel density estimate maps for each species and stressor, as well as maps of overlap for each species–stressor pair.

- **area_of_interest:** An `sf` vector object representing the polygon that defines the area of interest for the ByRA analysis. If not provided by the user, it is estimated by default based on the spatial distribution of the stressors.

First, let's check the summary stats!

```{r}
byra1$summary_stats
```

Now let’s visualize the main results using the *risaplot()* function. Since we are working with multiple species and stressors, the function will generate six plots.

The first two plots show the risk estimates (raw and reclassified) for each species, considering the isolated effect of each stressor. Plots 3 and 4 display the risk estimates (raw and reclassified) for each species when all stressors are combined. Finally, plots 5 and 6 present the ecosystem-level risk estimates (raw and reclassified), derived from the combined risk across all species.

```{r}
risaplot(byra1)
```

We can also visualize the maps that were considered for the risk estimates (KDE maps and their overlaps) using *risaplot()* applied to the *byra1$kde_maps* object. This will generate three plots: the first shows the KDE maps for each species, the second displays the KDE maps for each stressor, and the third presents the (reclassified) overlaps between each species–stressor pair.

```{r}
risaplot(byra1$kde_maps)
```








## References

Hines, EM, Johnson, A., Ponnampalam, L., Peter, C., Junchompoo, C., Vu, L., 
Thien, H., Caillat, M., and Verutes, G. (2020) Getting to the Bottom of Bycatch: 
A GIS-based toolbox to assess the risk of marine mammal bycatch. Endangered Species Research 
42:37-57 DOI: https://doi.org/10.3354/esr01037
